Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
proof again.

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{#1}}}
\newcommand{\cblue}[1]{{\color{blue}{#1}}}
\newcommand{\blankout}[1]{}

@@

While an ideal garbage collector is  impractical, in this work we show
that  static analysis  of  heap data  can help  practical garbage
collectors in reclaiming more  garbage.  The specific contributions of
this paper are
\begin{itemize}
\item  We formulate  a  liveness  analysis for  a  lazy first  order
  functional language. We show how  to compute function summaries in
  a  context  independent  manner  to achieve  a  context  sensitive
  analysis. We  further describe the transformation  of the liveness
  equations to grammar rules in a  way that the language accepted by
  the resulting grammar describes live heap locations.\mycomment{the
    proof of its correctness}
\item We prove  the undecidability of the grammar  that results from
  the  analysis.   To overcome  this,  we  present an  algorithm  to
  transform grammar  to finite  state automata  (DFA) such  that the
  language accepted by the  DFA over-approximates the liveness (and
  hence a safe approximation).
\item We have implemented a garbage  collector that can use the DFA
  to  retain live  cells that  also include  closures. We  have done
  extensive  experiments with  this garbage  collector and  obtained
  empirical results  that show  the effectiveness  of liveness-based
  garbage collection.
\end{itemize}

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
  contexts, we  shall use  the term 'expression' and the notation $e$  for
both  expressions and
  applications.}) are  given by transitions  of the form  $\rho, \stk,
\heap, e \rightarrow \rho', \stk', \heap', e'$.  Here \stk\ is a stack
of  continuation  frames.  Each  continuation  frame  is  of the  form
$(\rho, x, e)$, signifying that the  location of $x$ has to be updated
with the  value of the currently  evaluating expression, and $e$  is to be
evaluated  next  in  the  environment  $\rho$.   The  start  state  is
$([\;]_\rho,([\;]_\rho,  \ans,   (\print~\ans)):[\;]_{S}  ,  [\;]_{H},
(\mainpgm)$.  In  this, the initial  environment and the  initial heap
are  denoted  by  $[\;]_\rho$  and  $[\;]_H$, and  the  initial  stack
consists  of  a  single  continuation   frame  in  which  \ans\  is  a
distinguished  variable  that  would  be  updated with  the  value  of
(\mainpgm).  In  addition, \print\ is a function  modeling a printing
mechanism---a standard  run-time support assumption  for lazy languages
that prints the  value of (\mainpgm).  Notice the  use of the operator
$:$ for adding elements to the top of the stack.

@@

Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
evaluation path  terminating in  $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a point-wise-union of  the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
added  to the  union.  Also  notice a  consequences of  laziness---the
entire  expression including  the condition  is not  evaluated if  the
demand on it is $\emptyset$.  This  results in the empty map, which is
denoted by $\emptyset_{\cal{M}}$.

@@

\mycomment{The function  $\mathcal{L}$ now gives  the (total) liveness
  of  an   expression  $e$.   The  cases  $\SRETURN$   and  $\SIF$  are
  straightforward, but note the liveness $x.\epsilon$ generated by the
  latter.   The case  $(\LET\  z\leftarrow s\  \IN\  e')$ resembles  a
  three-address instruction:  the liveness of  $e$ is given  by taking
  the liveness, $\Lv$, of $e'$, killing any liveness of $z$ and adding
  any incremental  liveness from  $s$.  The main  subtlety is  how the
  liveness of  $z$ in $\Lv$  is converted to  a demand $\Lv(z)$  to be
  placed on $s$ via $\mathit{ref}(s,\Lv(z),\Lfonly)$.

We make three observations: firstly the rule ({\sc live-define}) has a
least  solution  as  $\mathcal{L}(\cdot)$  is monotonic  in  $\sigma$;
secondly  that  ({\sc  live-define})   resembles  the  rule  for  type
inference of mutually recursive  function definitions, and thirdly the
asymmetry of  demand and liveness (compared to  post- and pre-liveness
classically) is due to the functional formulation here.
}

@@

We shall now  outline a proof of correctness  of the liveness analysis
presented  in  Section~\ref{sec:liveness-analysis}.   This requires  a
significant modification  to the technique  used in \cite{asati14lgc}.
\cblue{Good points to  make in the main text, right  in the small step
  semantics  section: Note  that the  state of  the  transition system
  consists of the current evaluation context, given by $\rho$ and $e$,
  and  the suspended evaluation  contexts recorded  in the  stack $S$.
  The heap $H$  is common to all contexts.}

@@

\section{Computing liveness and its encoding as DFA}\label{sec:computing}
Section~\ref{sec:liveness} gave  a context-sensitive liveness analysis%
  \mycomment{\color  {red}  and proved  it  correct  with  reference  to  a  {\em
    minefield} semantics}.  In  particular, $\mathcal{L}$ and \pmap{f}
together described a liveness set for each program point in a function
in terms of  a given $\sigma$ and \Lfonly.  However,  we still have to
describe how to obtain demand transformers \Lfonly\ from the rule {\sc
  live-define} and how to compute the specific demand $\sigma$ on each
function.
%%
To  answer these  questions, it  is convenient  to cast  the equations
arising  out of  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
grammar.   To do  so, we  need  to modify  the rules  themselves to  a
different form.

@@

\appendix

\section{Minefield semantics for the language}

>>>>

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
%\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
%\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
%\newcommand{\cblue}[1]{{\color{blue}{#1}}}
%\newcommand{\blankout}[1]{}

@@

While an ideal garbage collector is  impractical, in this work we show
that  static analysis  of  heap data  can help  practical garbage
collectors in reclaiming more  garbage.  The specific contributions of
this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language  and prove  of its correctness.  We show  how to
  compute  function  summaries  in  a context  independent  manner  to
  achieve  a  context  sensitive  analysis. We  further  describe  the
  transformation of the  liveness equations to grammar rules  in a way
  that the language  accepted by the resulting  grammar describes live
  heap locations.
\item We prove  the undecidability of the grammar  that results from
  the  analysis.   To overcome  this,  we  present an  algorithm  to
  transform grammar  to finite  state automata  (DFA) such  that the
  language accepted by the  DFA over-approximates the liveness (and
  hence a safe approximation).
\item We have implemented a garbage  collector that can use the DFA
  to  retain live  cells that  also include  closures. We  have done
  extensive  experiments with  this garbage  collector and  obtained
  empirical results  that show  the effectiveness  of liveness-based
  garbage collection.
\end{itemize}

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
  contexts, we  shall use the  term 'expression' and the  notation $e$
  for both expressions and applications.}) are given by transitions of
the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
Here \stk\ is a stack of continuation frames.  Each continuation frame
is of the form $(\rho, x, e)$, signifying that the location of $x$ has
to be updated  with the value of the  currently evaluating expression,
and $e$ is to be evaluated  next in the environment $\rho$.  Note that
the state of the transition  system consists of the current evaluation
context,  given  by  $\rho$  and $e$,  and  the  suspended  evaluation
contexts recorded  in the stack  $S$.  The heap  $H$ is common  to all
contexts.

The     start      state     is      $([\;]_\rho,([\;]_\rho,     \ans,
(\print~\ans)):[\;]_{S} , [\;]_{H}, (\mainpgm)$.  In this, the initial
environment  and  the initial  heap  are  denoted by  $[\;]_\rho$  and
$[\;]_H$,  and the  initial stack  consists of  a single  continuation
frame in which \ans\ is a distinguished variable that would be updated
with  the value  of (\mainpgm).   In addition,  \print\ is  a function
modeling a printing mechanism---a standard run-time support assumption
for lazy  languages that prints  the value of (\mainpgm).   Notice the
use of the operator $:$ for adding elements to the top of the stack.

@@

Consider the  $\mathcal{L}$-rules for {\LET}, {\SIF},  and {\SRETURN}.
The expression  $(\SRETURN~\psi:x)$ with  a demand $\sigma$  forces an
evaluation  of  $x$ and  hence  a  new map  $[\mathit{\epath}  \mapsto
  \{x.\sigma\}]  $ needs  to be  created,  where $\epath$  is a  fresh
evaluation path  terminating in  $\psi$.  The  map for  the expression
$(\SIF~\psi:x~e_1~e_2)$ is a point-wise-union of the maps of $e_1$ and
$e_2$. In addition, since the condition also triggers an evaluation of
$x$, the map $[\mathit{\epath} \mapsto \{x.\epsilon\}]$ is created and
added  to the  union.   Also notice  a  consequence of  laziness---the
entire  expression including  the condition  is not  evaluated if  the
demand on it is $\emptyset$.  This  results in the empty map, which is
denoted by $\emptyset_{\cal{M}}$.

@@

@@

We shall now  outline a proof of correctness of  the liveness analysis
presented   in    Section~\ref{sec:liveness-analysis}.    This   proof
technique  is  similar to  the  technique  used in  \cite{asati14lgc}.

@@

\section{Computing liveness and its encoding as DFA}\label{sec:computing}
Section~\ref{sec:liveness} gave a  context-sensitive liveness analysis
and proved it  correct with reference to a  {\em minefield} semantics.
In  our  analysis, $\mathcal{L}$  and  \pmap{f}  together described  a
liveness set for each program point in  a function in terms of a given
$\sigma$  and \Lfonly.   However, we  still  have to  describe how  to
obtain demand  transformers \Lfonly\  from the rule  {\sc live-define}
and how to compute the specific demand $\sigma$ on each function.
%%
To  answer these  questions, it  is convenient  to cast  the equations
arising  out of  $\mathit{ref}$ and  $\mathcal{L}$ as  the rules  of a
grammar.   To do  so, we  need  to modify  the rules  themselves to  a
different form.

@@

\appendix
\newcommand{\cred}[1]{{\color{red}{#1}}}
\section{Minefield semantics for the language}
